---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  md_document: default
  html_notebook: default
---

 Stock Price and Volatility Analysis (Notebook)
---
title: "Stock Price and Volatility Analysis"
output: html_document
---

## 1. Setup

```{r, warning=FALSE, message=FALSE, error=FALSE}
# your code here
library(tidyverse)
library(lubridate)
library(tseries)
library(FinTS)
library(zoo)
library(lmtest)
library(stats4)
```

---

## 2. Load and Inspect Data

```{r}
data <- read.csv("C:/Users/admin/Desktop/New folder/su work/1.2/timeseries/stock.csv")

head(data)
str(data)

colSums(is.na(data))
```

We focus only on the relevant variables: date and adjusted closing price.

```{r}
data1 <- data[, c("Date", "Adj..Close")]
colnames(data1) <- c("Date", "Price")
```

---

## 3. Data Preparation

Convert dates and sort the dataset chronologically.

```{r}
data1$Date <- mdy(trimws(data1$Date))
data2 <- data1 %>% arrange(Date)

head(data2)
```

---

## 4. Price Visualization

```{r}
plot(data2$Date, data2$Price, type = "l",
     main = "Stock Price Over Time",
     xlab = "Date", ylab = "Price")
grid()
```

The price series shows long-term trends but does not directly reveal volatility patterns.

---

## 5. Compute Log Returns

```{r}
data2$log_return <- c(NA, diff(log(data2$Price)))
data2 <- data2[-1, ]

head(data2)
```

```{r, warning=FALSE, message=FALSE, error=FALSE}
# your code here
plot(data2$Date, data2$log_return, type = "l",
     col = "green", lwd = 2,
     main = "Log Returns",
     xlab = "Date", ylab = "Log Return")
abline(h = 0, col = "red", lty = 2)
```

Log returns fluctuate around zero and highlight volatility clustering.

---

## 6. Statistical Properties

### Stationarity

```{r}
adf.test(data2$log_return)
```

### Autocorrelation

```{r}
acf(data2$log_return)
pacf(data2$log_return)

Box.test(data2$log_return, lag = 20, type = "Ljung-Box")
```

### Heteroskedasticity

```{r}
ArchTest(data2$log_return)
```

These tests help determine whether volatility models are appropriate.

---

## 7. Volatility Estimation

### Overall Volatility

```{r}
sd(data2$log_return)
```

### Rolling Volatility (20 days)

```{r}
data2$vol_20 <- rollapply(data2$log_return, 20, sd, fill = NA)

plot(data2$Date, data2$vol_20, type = "l",
     main = "20-Day Rolling Volatility",
     xlab = "Date", ylab = "Volatility")
```

### Rolling Volatility (5 days)

```{r}
data2$vol_5 <- rollapply(data2$log_return, 5, sd, fill = NA)

plot(data2$Date, data2$vol_5, type = "l",
     main = "5-Day Rolling Volatility",
     xlab = "Date", ylab = "Volatility")
```

Volatility clearly changes over time, confirming non-constant variance.

---

## 8. Risk-Return Relationship (OLS)

```{r}
ols_model <- lm(log_return ~ vol_5, data = data2)
summary(ols_model)
```

```{r}
ols_model2 <- lm(log_return ~ vol_20, data = data2)
summary(ols_model2)
```

These models explore whether higher volatility is associated with higher returns.

---

## 9. Model Diagnostics

```{r}
# Residual mean
mean(residuals(ols_model))

# Homoscedasticity
plot(ols_model$fitted.values, residuals(ols_model))
abline(h = 0, col = "red")

bptest(ols_model)

# Normality
qqnorm(residuals(ols_model))
qqline(residuals(ols_model), col = "red")

shapiro.test(residuals(ols_model))

# Autocorrelation
dwtest(ols_model)
```

---

## 10. Maximum Likelihood Estimation

Prepare clean data:

```{r}
data_mle <- na.omit(data2[, c("log_return", "vol_5")])
```

Define likelihood function:

```{r}
loglik <- function(alpha, beta, sigma) {
  mu <- alpha + beta * data_mle$vol_5
  -sum(dnorm(data_mle$log_return, mean = mu, sd = sigma, log = TRUE))
}
```

Estimate parameters:

```{r, warning=FALSE, message=FALSE, error=FALSE}
mle_model <- mle(loglik,
                 start = list(alpha = 0, beta = 0.1, sigma = 0.02),
                 method = "BFGS")

summary(mle_model)
```

---

## 11. Key Takeaways

* Returns are stationary but exhibit volatility clustering
* ARCH effects suggest time-varying variance
* Rolling volatility confirms changing risk levels
* OLS provides basic insights but has limitations
* MLE improves parameter estimation

